---
output: 
  html_document: 
    # css: style1.css
---

**Generate Sentence annotators using 'Stanford CoreNLP'**  
Stanford CoreNLP provides a set of natural language analysis tools including POS Tagger , Named entity recognition , corefernce , dependencies , Sentiment and and Parser.
  
```{r, echo=TRUE, warning=FALSE, results='markup', split=TRUE}
##install.packages("coreNLP")
library(coreNLP)
```

```{r, echo=TRUE, warning=FALSE}
##Annotate a string of text
#downloadCoreNLP()
initCoreNLP()
s <- paste(c( "Mr.and Mrs. Dursley, of number four, Privet ",
              "Drive, were proud that they were ",
              "perfectly normal,thank you very much.\n",    
              "They were the last people you'd expect to be" ,
              "involved in anything strange or mysterious, " ,
              "because they just didn't hold with such nonsense."),
           collapse = "" )

s                  

anObj<- annotateString(s)
anObj
```
There are two sentences in this annotation object anObj and the number of tokens are 49.   

Annotation of first two lines of Albert Camus' L'Etranger
```{r, echo=TRUE, warning=FALSE}
annoEtranger
```
There are two sentences in the object annoEtranger and the number of tokens are 16.   

Annotation of first line of JK Rowling's The Philosopher's Stone
```{r}
annoHp
```
There is a single sentence in the object named annoHP and the number of tokens is 27.


**Tokens as data frame**  
The function 'getToken' in coreNLP gives the tokens along with the POS and NER values in a dataframe. Lemma generates the base form of all the tokens in the sentence. 
```{r, echo=TRUE, warning=FALSE}
tokenEr<-getToken(annoEtranger)
tokenEr
tokenHP<-getToken(annoHp)
tokenHP
```


**POS Tagger**  
Parts of speech tagger assigns the parts of speech tag to each token in the text.

|Number  |Tag  |Description
---|---|---
|1.      | CC	 |Coordinating conjunction
|2.	     | CD	 |Cardinal number
|3.	     | DT	 |Determiner
|4.	     | EX	 |Existential there
|5.	     | FW	 |Foreign word
|6.	     |IN	 |Preposition or subordinating conjunction
|7.	     |JJ	 |Adjective
|8.	     |JJR	 |Adjective, comparative
|9.	     |JJS	 |Adjective, superlative
|10.     |LS   |List item marker
|11.     |MD   |Modal
|12.     |NN   |Noun, singular or mass
|13.     |NNS  |Noun, plural
|14.     |NNP	 |Proper noun, singular
|15.	   |NNPS |Proper noun, plural
|16.	   |PDT	|Predeterminer
|17.	    |POS	|Possessive ending
|18.	    |PRP	|Personal pronoun
|19.	    |PRP$	|Possessive pronoun
|20.	    |RB	|Adverb
|21.	    |RBR	|Adverb, comparative
|22.	    |RBS	|Adverb, superlative
|23.	    |RP	|Particle
|24.	    |SYM	|Symbol
|25.	    |TO	|to
|26.	    |UH	|Interjection
|27.	    |VB	|Verb, base form
|28.	    |VBD	|Verb, past tense
|29.	    |VBG	|Verb, gerund or present participle
|30.	    |VBN	|Verb, past participle
|31.	    |VBP	|Verb, non-3rd person singular present
|32.	    |VBZ	|Verb, 3rd person singular present
|33.	    |WDT	|Wh-determiner
|34.	    |WP	|Wh-pronoun
|35.	    |WP$	|Possessive wh-pronoun
|36.	    |WRB	|Wh-adverb

**Tabulated tokens with Parts Of Speech**
```{r}
table(tokenEr$POS)
```
```{r}
table(tokenHP$POS)
```

**Tabulated tokens with Named Entity Recognition**  
NER recognizes named (Person, Location,Organization, MISC) and numerical(Money,Number,Date,Time,Duration,Set) entities.
```{r}
table(tokenEr$NER)
```
```{r}
table(tokenHP$NER)
```

**Getting coreference**    
Coreference helps in finding all the expressions that refer to a particular entity in the given text.
```{r, echo=TRUE}

getCoreference(anObj)
```

```{r, echo=TRUE, warning=FALSE}
getCoreference(annoHp)
```
**Get Dependencies**  
Dependencies are binary relations which give the grammatical relations between a 'governor' and a 'dependent'.
```{r, echo=TRUE, warning=FALSE}
getDependency(annoEtranger)
getDependency(annoHp)
```
'Sentence' gives the id of the sentence followed by the 'governor' and the 'dependent' words.'Type' indicates the grammatical relationship between the two.'Governorindex' and 'dependentidx' represent the word position in the sentence.

**OpenIE to extract the relationship in a sentence**
```{r, echo=TRUE}
getOpenIE(annoHp)
```


**Getting a parse tree**    
Parse tree helps in identifying the grammatical relationships between words in a sentence.Mostly used for visualization.
```{r, echo=TRUE, warning=FALSE}
getParse(annoEtranger)
```
```{r}
getParse(annoHp)
```

**Sentiment scores**  
Sentiment analysis is performed over the parse tree and each node is assigned a sentiment score based on which the overall sentiment of the given text is determined.
```{r, echo=TRUE, warning=FALSE}
getSentiment(annoEtranger)
getSentiment(annoHp)
```
The sentiment for annotation object annoEtranger is negative in the first sentence and neutral in the second sentence.

The sentiment for annotation object annoHP which is the first line of Harry potter and the philosopher's stone is Positive.
